### 2022-11-17 chapter1/chapter1.tex
Modified chapter1 to be a subfile and fixed some minor formatting issues

\documentclass[11pt,letterpaper]{article}

% PACKAGE CONFIGURATIONS
\usepackage{amsmath}

\usepackage{graphicx}
\graphicspath{{./figures/}}

\usepackage[margin=1in]{geometry}
%\usepackage{setspace}  % double spacing
%\doublespacing
\usepackage[skip=10pt]{parskip}

\usepackage{hyperref}

\usepackage[doi=false,url=false]{biblatex}
\addbibresource{refs.bib}

\usepackage[labelfont=bf, labelsep=period]{caption}
\renewcommand{\figurename}{Fig}

\begin{document}

% Title must be 250 characters or less
\begin{flushleft}
{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
\newline
\\
Marc Singleton\textsuperscript{1},
Michael Eisen\textsuperscript{1,2,3,*}
\\
\bigskip
\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
\\
\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
\\
\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
\\
\bigskip

* mbeisen@berkeley.edu

\end{flushleft}

% Abstract must be 300 words or less
\begin{abstract}
    \noindent Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
\end{abstract}

@@

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: --globalpair --maxiterate 1000 --thread 1 --anysymbol --allowshift --unalignlevel 0.4 --leavegappyregion~\cite{Katoh2013}. Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment. Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

@@

\end{document}

>>>>

% Abstract must be 300 words or less
\begin{abstract}
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
\end{abstract}

@@

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}. Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment. Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

@@
